---
title: "Event Sequence Visualization Case Study (R / TraMineR)"
subtitle: "Reference visual outputs for side-by-side comparison with Sequenzo Python"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
---

## Goal

This tutorial demonstrates the key **event-sequence visualizations** in the original R ecosystem:

- `seqpcplot()` / `plot(seqelist, type = "pc")`
- `plot.subseqelist()`
- `plot.subseqelistchisq()`
- `seqedplot(type = "survival" | "hazard")`

The purpose is to produce a clear R baseline so that your audience can compare it directly with the Python notebook in the same folder.

## 1) Load packages and data

```{r}
#| label: setup
#| eval: true
suppressPackageStartupMessages({
  library(TraMineR)
  library(TraMineRextras)
})

# Use built-in TraMineR example datasets
data(actcal)      # state-sequence style table
data(actcal.tse)  # time-stamped event format

cat("Rows in actcal:", nrow(actcal), "\n")
cat("Rows in actcal.tse:", nrow(actcal.tse), "\n")
head(actcal.tse)
```

## 2) Build state sequence and event sequence objects

```{r}
#| label: build-sequences
#| eval: true
# State sequences from monthly activity columns
actcal.seq <- seqdef(actcal[, 13:24])

# Transition matrix and event sequence object
transition <- seqetm(actcal.seq, method = "transition")
transition[1, 1:4] <- c("FullTime", "Decrease,PartTime", "Decrease,LowPartTime", "Stop")
transition[2, 1:4] <- c("Increase,FullTime", "PartTime", "Decrease,LowPartTime", "Stop")
transition[3, 1:4] <- c("Increase,FullTime", "Increase,PartTime", "LowPartTime", "Stop")
transition[4, 1:4] <- c("Start,FullTime", "Start,PartTime", "Start,LowPartTime", "NoActivity")

actcal.eseq <- seqecreate(actcal.seq, event_representation = transition)
actcal.eseq
alphabet(actcal.eseq)
```

## 3) Frequent subsequences and discriminating subsequences

```{r}
#| label: frequent-subseq
#| eval: true
# Frequent subsequences with support >= 1%
fsub <- seqefsub(actcal.eseq, pmin.support = 0.01)
fsub
head(fsub$data, 10)

# Group comparison by sex (Bonferroni-corrected chi-square)
discr <- seqecmpgroup(fsub, group = actcal$sex, method = "bonferroni")
discr
head(discr$data, 10)
```

## 4) Visualization A — Parallel coordinates (`seqpcplot`)

```{r}
#| label: plot-seqpc
#| eval: true
#| fig-cap: "TraMineR event-sequence parallel coordinates (overall and by sex)."
par(mfrow = c(1, 2))
seqpcplot(actcal.eseq, main = "seqpcplot: all sequences")
# seqpcplot(actcal.eseq, group = actcal$sex, main = "seqpcplot: grouped by sex")
par(mfrow = c(1, 1))
```

## 5) Visualization B — Frequent subsequence bars (`plot.subseqelist`)

```{r}
#| label: plot-subseqelist
#| eval: true
#| fig-cap: "Top frequent subsequences by support and by count."
par(mfrow = c(1, 2))
plot(fsub[1:12], main = "plot.subseqelist (Support)")
plot(fsub[1:12], freq = fsub$data$Count[1:12], main = "plot.subseqelist (Count)")
par(mfrow = c(1, 1))
```

## 6) Visualization C — Group discriminating subsequences (`plot.subseqelistchisq`)

```{r}
#| label: plot-subseqelistchisq
#| eval: true
#| fig-cap: "Discriminating subsequences: frequency and Pearson residual views."
par(mfrow = c(1, 2))
plot(discr[1:10], ptype = "freq")
plot(discr[1:10], ptype = "resid")
par(mfrow = c(1, 1))
```

## 7) Visualization D — Event survival/hazard (`seqedplot`)

```{r}
#| label: plot-seqed
#| eval: true
#| fig-cap: "Event-sequence survival and hazard visualizations from TraMineRextras."
# Keep first 200 ids for fast rendering and cleaner visuals
actcal.tse.small <- actcal.tse[1:200, ]
iseq <- unique(actcal.tse.small$id)
actcal.small <- actcal[rownames(actcal) %in% iseq, ]

actcal.seqe.small <- seqecreate(actcal.tse.small)
seqelength(actcal.seqe.small) <- rep(12, length(iseq))

par(mfrow = c(1, 2))
seqedplot(actcal.seqe.small, type = "survival", group = actcal.small$sex, lwd = 2)
seqedplot(actcal.seqe.small, type = "hazard", breaks = 6, group = actcal.small$sex, lwd = 2)
par(mfrow = c(1, 1))
```

## Notes for comparison

- This file is the **R reference**.
- The Python counterpart in this folder reproduces the same four visualization families with Sequenzo.
- For fair comparison in your presentation, keep:
  - same subset size,
  - same grouping variable,
  - similar color palette and figure size.
